---
title: "Population Scale Analysis"
author: "Angela Liu"
format: gfm
---

We will be taking a look at genetic differences at a population level and see if there's an association between the asthma related SNPs on ORMDL3.


>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.


Let's see how many samples we have:

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```



```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
summary(expr)
```

```{r}
# save the boxplot information into variable medExp
medExp <- boxplot(exp~geno, data = expr)
medExp$stats
medExp$stats[3,]
```


The sample size for A/A genotype is 108. A/G has 233 samples and G/G has 121 samples.

The third row of the stats of the boxplot reflect the median of each genotype. The medians go as followed: 31.25 for A/A, 25.06 for A/G, 20.07 for G/G. 


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```


Let's make a boxplot of our data:
```{r}
# notch adds a belt
ggplot(expr) + aes(x=geno, y=exp, fill = geno) +
  geom_boxplot(noth=TRUE)
```


From the plot, we can see that A/A has a higher expression value than G/G. Yes, the SNP affects the expression of ORMDL3 as the genotype for G/G is reduced with its lower median. 

